---
title: "GSS_Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GSS_Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## This workflow
This workflow will take you through using GSS on "patient data" to identify progression towards Tcell exhaustion. 

### Necessary Packages for this Example Workflow: 
```{r, eval=FALSE}
library(GeneSwitchesScorer)
library(Seurat)
library(GeneSwitches)
library(slingshot)

library(SingleCellExperiment)
# library(SeuratObject)
# library(sp)
library(RColorBrewer)
# library(data.table)
# library(reticulate)
# library(patchwork)
# library(parallel)
# library(ggplot2)
# library(Matrix)
```

## Input Datasets
This workflow uses data from "A Single-Cell Tumor Immune Atlas for Precision Oncology" \
**TODO REF**:\
https://zenodo.org/record/5205544 \
It also uses **###** as the sample/patient data. 
both are seurat objects.

Download these from the above links and load them into your environment using:
```{r, eval = FALSE}
atlas.seu <- readRDS("~/R Packages/mini_data/mini_TICAtlas.rds")
pre.seu <- readRDS("~/R Packages/mini_data/mini_PTCL_PRE.rds")
post.seu <- readRDS("~/R Packages/mini_data/mini_PTCL_POST.rds")
```

### Optional: view the umaps of the input data. 
```{r, eval = FALSE}
DimPlot(object = pre.seu, reduction = "umap", group.by = "celltype" ,label = T)
DimPlot(object = post.seu, reduction = "umap", group.by = "celltype" ,label = T)
DimPlot(object = atlas.seu, reduction = "umap", group.by = "lv1_annot" ,label = T)
```

## Subset the objects to only be the trajectory that we are interested in. 
```{r, eval = FALSE}
# subsetting the pre treatment patient data to only include the exhaustion trajectory.
pre_T.seu <- subset(x = pre.seu, subset = celltype %in% c("Early_Tcells/CD3+CD4+TIGIT+GATA3+",
                                                          "CD3+CD4-CD8-TIGIT+_Tcell3",
                                                          "CD3+CD4-CD8+TIGIT+_Tcell2",
                                                          "CD3+CD4-CD8+TIGIT+_Tcells",
                                                          "Cytotoxic_CD8Tcell",
                                                          "CD3+TIGIT+Tcell",
                                                          "Tcell1"))

# subsetting the post treatment patient data to only include the exhaustion trajectory.
post_T.seu <- subset(x = post.seu, subset = celltype %in% c("CD3+PDCD1+TIGIT+Activated",
                                                            "Tcell2_CD3+GATA3+TIGIT+",
                                                            "Tcell_CD3+GATA3+TIGIT+",
                                                            "Cytotoxic_CD8Tcell",
                                                            "Tcell_CD3+TIGIT+",
                                                            "Activated_Tcell",
                                                            "Tcell2"))

# Subsetting the Atlas data into 3 types of Tcells for the exhaustion trajectory.
atlas_exhaustion.seu <- subset(x = atlas.seu, subset = lv1_annot %in% c("T cells naive",
                                                                        "CD8 cytotoxic", 
                                                                        "CD8 terminally exhausted"))
```

### Optional: view the umaps of the subsetted data. 
```{r, eval = FALSE}
DimPlot(object = pre_T.seu, reduction = "umap", group.by = "celltype" ,label = T)
DimPlot(object = post_T.seu, reduction = "umap", group.by = "celltype" ,label = T)
DimPlot(object = atlas_exhaustion.seu, reduction = "umap", group.by = "lv1_annot" ,label = T)
```


## Process the atlas data using Seurat. 
```{r, eval = FALSE}
# Scale Data
atlas_exhaustion.seu <- ScaleData(atlas_exhaustion.seu)
# Run PCA
atlas_exhaustion.seu <- RunPCA(atlas_exhaustion.seu)
# Run UMAP
atlas_exhaustion.seu <- RunUMAP(atlas_exhaustion.seu, dims = 1:30)
# FindNeighbors
atlas_exhaustion.seu <- FindNeighbors(atlas_exhaustion.seu, reduction = "pca", dims = 1:30)
# FindClusters
atlas_exhaustion.seu <- FindClusters(atlas_exhaustion.seu, resolution = 0.5)
```

## Convert the Seurat objects to SingleCellExperiment objects.
```{r, eval = FALSE}
# converting the pre-treatment patient data to SCE.
pre.sce <- as.SingleCellExperiment(pre.seu)

# converting the pre-treatment patient data to SCE.
post.sce <- as.SingleCellExperiment(post.seu)

# converting the atlas data to SCE.
atlas_exhaustion.sce <- as.SingleCellExperiment(atlas_exhaustion.seu)
```

## Run slingshot on the atlas data
```{r, eval = FALSE}
atlas_exhaustion.sce  <- slingshot(atlas_exhaustion.sce, 
                                   clusterLabels = "lv1_annot", 
                                   start.clus  = "T cells naive", 
                                   end.clus = "CD8 terminally exhausted", 
                                   reducedDim = "UMAP")
```

#### Check:
```{r, eval = FALSE}
summary(atlas_exhaustion.sce$slingPseudotime_1)
```


### Plot the trajectory
```{r, eval = FALSE}
colors <- colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)
plotcol <- colors[cut(atlas_exhaustion.sce$slingPseudotime_1, breaks=100)]
plot(reducedDims(atlas_exhaustion.sce)$UMAP, col = plotcol, pch=16, asp = 1)
lines(SlingshotDataSet(atlas_exhaustion.sce), lwd=2, col='black')
```

## Convert the SingleCellExperiment objects to GeneSwitches Objects.
```{r, eval = FALSE}
# convert the pre treatment patient data to a GS obj. 
gs_pre <- SingleCellExperiment(assays = List(expdata = logcounts(pre.sce)))
# convert the post treatment patient data to a GS obj. 
gs_post <- SingleCellExperiment(assays = List(expdata = logcounts(post.sce)))
# convert the atlas data to a GS obj.
gs_atlas <- SingleCellExperiment(assays = List(expdata = logcounts(atlas_exhaustion.sce)))
colData(gs_atlas)$Pseudotime <- -atlas_exhaustion.sce$slingPseudotime_1
reducedDims(gs_atlas) <- SimpleList(UMAP = reducedDim(atlas_exhaustion.sce, "UMAP", withDimnames=TRUE))
```

## Binarize
When checking if the 0.2 cutoff is appropriate I found there to be negative expression... \
```{r, eval = FALSE} hist(as.numeric(assays(gs_atlas)$expdata))``` \
```{r, eval = FALSE}
bn_cutoff <- 0.3
# binarize pre treatment patient data
gs_pre<- binarize_exp(gs_pre, fix_cutoff = TRUE, binarize_cutoff = bn_cutoff)
# binarize post treatment patient data
gs_post<- binarize_exp(gs_post, fix_cutoff = TRUE, binarize_cutoff = bn_cutoff)
# binarize atlas data
gs_atlas_binerized <- binarize_exp(gs_atlas, fix_cutoff = TRUE, binarize_cutoff = bn_cutoff)
```

## fit logistic regression and find the switching pseudo-time point for each gene
```{r, eval = FALSE}
gs_atlas_glm <- find_switch_logistic_fastglm(gs_atlas_binerized, downsample = FALSE, show_warning = FALSE)
```

## Filter Switching Genes
### ~*currently using the default cutoff of 0.03 which gives 29 genes.*~
### ~*currently using a cutoff of 0.01 which gives 108 genes.*~
### *currently using a cutoff of 0.005 which gives 148 genes.*
```{r, eval = FALSE}
gs_atlas_r2005 <- filter_switchgenes(gs_atlas_glm, allgenes = TRUE,r2cutoff = 0.005)
```


### Optional: view the timeline of the filtered switching genes from the Atlas.
### *Seems a bit odd, maybe it is to do with using logged expression data*
```{r, eval = FALSE}
plot_timeline_ggplot(gs_atlas_r2005, timedata = colData(gs_atlas)$Pseudotime, txtsize = 3)
```

## Select evenly distributed switching genes
remember to define the minimum time spacing
```{r, eval = FALSE}
gs_scorer_genes <- select_evenly_distributed_switching_genes(gs_atlas_r2005, min_time_spacing = 5)
```
